Latentcor: An R Package for Latent Correlation Estimation

### 2021-06-03 paper/paper.Rmd
Fixed

Compare to MixedCCA, standalone, new functionality, memory footprint.

@@

*Definition 1* Fan et al. (2017) considered the problem of estimating $\Sigma$ for the latent Gaussian copula model based on Kendall's $\tau$. Given the observed data $(X_{1j}, X_{1k}), ..., (X_{nj}, X_{nk})$ for variables $X_{j}$ and $X_{k}$, Kendall's $\tau$ is defined as
$$
\hat{\tau}_{jk}=\frac{2}{n(n-1)}\sum_{1\leq i <i'\leq n} sign(X_{ij}-X_{i'j})sign(X_{ik}-X_{i'k})
$$
*Theorem 1* Let $W_{1}\in\cal{R}^{p_1}$, $W_{2}\in\cal{R}^{p_2}$, $W_{3}\in\cal{R}^{p_3}$ be such that $W=(W_{1}, W_{2}, W_{3})\sim NPN(0,\Sigma,f)$ with $p=p_{1}+p_{2}+p_{3}$. Let $X=(X_{1}, X_{2}, X_{3})\in\cal{R}^{p}$ satisfy $X_{j}=W_{j}$ for $j=1,...,p_{1}$, $X_{j}=I(W_{j}>c_{j})$ for $j=p_{1}+1,...,p_{1}+p_{2}$ and $X_{j}=I(W_{j}>c_{j})W_{j}$ for $j=p_{1}+p_{2}+1,...,p$ with $\Delta_{j}=f(c_{j})$. The rank-based estimator of $\Sigma$ based on the observed $n$ realizations of $X$ is the matrix $\hat{R}$ with $\hat{r}_{jj}=1$, $\hat{r}_{jk}=\hat{r}_{kj}=F^{-1}(\hat{\tau}_{jk})$ with block structure
$$
\hat{R}=\left(\begin{array}\\
F^{-1}_{CC}(\hat{\tau})\hspace{.2in} F^{-1}_{CB}(\hat{\tau})\hspace{.2in} F^{-1}_{CT}(\hat{\tau})\\
F^{-1}_{BC}(\hat{\tau})\hspace{.2in} F^{-1}_{BB}(\hat{\tau})\hspace{.2in} F^{-1}_{BT}(\hat{\tau})\\
F^{-1}_{TC}(\hat{\tau})\hspace{.2in} F^{-1}_{TB}(\hat{\tau})\hspace{.2in} F^{-1}_{TT}(\hat{\tau})
\end{array}\right)
$$
The original method is taking estimated Kendall's $\hat{\tau}$ and other parameters to calculate latent correlation $\hat{r}$. Whereas the approximated method is using multilinear interpolation to approximate latent correlation $\hat{r}$ via pre-calculated grid values [@yoon2021fast].

@@

 Table to show memory improvement compare to mixedCCA.
 Generate data for pairs.
 One example of original call, one for interpolation call.

>>>>

Compare to MixedCCA, memory footprint.

@@

*Definition 1* Fan et al. (2017) considered the problem of estimating $\Sigma$ for the latent Gaussian copula model based on Kendall's $\tau$. Given the observed data $(X_{1j}, X_{1k}), ..., (X_{nj}, X_{nk})$ for variables $X_{j}$ and $X_{k}$, Kendall's $\tau$ is defined as
$$
\hat{\tau}_{jk}=\frac{2}{n(n-1)}\sum_{1\leq i <i'\leq n} sign(X_{ij}-X_{i'j})sign(X_{ik}-X_{i'k})
$$
*Theorem 1* Let $W_{1}\in\cal{R}^{p_1}$, $W_{2}\in\cal{R}^{p_2}$, $W_{3}\in\cal{R}^{p_3}$, $W_{4}\in\cal{R}^{p_4}$ be such that $W=(W_{1}, W_{2}, W_{3}, W_{4})\sim NPN(0,\Sigma,f)$ with $p=p_{1}+p_{2}+p_{3}+p_{4}$. Let $X=(X_{1}, X_{2}, X_{3}, X_{4})\in\cal{R}^{p}$ satisfy $X_{j}=W_{j}$ for $j=1,...,p_{1}$, $X_{j}=I(W_{j}>c_{j})$ for $j=p_{1}+1,...,p_{1}+p_{2}$, $X_{j}=I(W_{j}>c_{j})W_{j}$ for $j=p_{1}+p_{2}+1,...,p$ and $X_{j}=I(W_{j}>c_{j}^{1})+I(W_{j}>c_{j}^{2})$ with $\Delta_{j}=f(c_{j})$, $\Delta_{j}^{1}=f(c_{j}^{1})$ and $\Delta_{j}^{2}=f(c_{j}^{2})$. The rank-based estimator of $\Sigma$ based on the observed $n$ realizations of $X$ is the matrix $\hat{R}$ with $\hat{r}_{jj}=1$, $\hat{r}_{jk}=\hat{r}_{kj}=F^{-1}(\hat{\tau}_{jk})$ with block structure
$$
\hat{R}=\left(\begin{array}\\
F^{-1}_{CC}(\hat{\tau})\hspace{.2in} F^{-1}_{CB}(\hat{\tau})\hspace{.2in} F^{-1}_{CT}(\hat{\tau})\hspace{.2in} F^{-1}_{CN}(\hat{\tau})\\
F^{-1}_{BC}(\hat{\tau})\hspace{.2in} F^{-1}_{BB}(\hat{\tau})\hspace{.2in} F^{-1}_{BT}(\hat{\tau})\hspace{.2in} F^{-1}_{BN}(\hat{\tau})\\
F^{-1}_{TC}(\hat{\tau})\hspace{.2in} F^{-1}_{TB}(\hat{\tau})\hspace{.2in} F^{-1}_{TT}(\hat{\tau})\hspace{.2in} F^{-1}_{TN}(\hat{\tau})\\
F^{-1}_{NC}(\hat{\tau})\hspace{.2in} F^{-1}_{NB}(\hat{\tau})\hspace{.2in} F^{-1}_{NT}(\hat{\tau})\hspace{.2in} F^{-1}_{NN}(\hat{\tau})
\end{array}\right)
$$
The original method is taking estimated Kendall's $\hat{\tau}$ and other parameters to calculate latent correlation $\hat{r}$. Whereas the approximated method is using multilinear interpolation to approximate latent correlation $\hat{r}$ via pre-calculated grid values [@yoon2021fast].

@@

 Table to show memory improvement compare to mixedCCA.
```{r}
library(latentcor)
### Data setting
n <- 1000; p1 <- 1; p2 <- 1 # sample size and dimensions for two datasets.
maxcancor <- 0.9 # true canonical correlation

### Correlation structure within each data set
set.seed(0)
perm1 <- sample(1:p1, size = p1);
Sigma1 <- autocor(p1, 0.7)[perm1, perm1]
blockind <- sample(1:3, size = p2, replace = TRUE);
Sigma2 <- blockcor(blockind, 0.7)
mu <- rbinom(p1+p2, 1, 0.5)

### true variable indices for each dataset
trueidx1 <- 1
trueidx2 <- 1

# Data generation
simdata <- GenerateData(n=n, trueidx1 = trueidx1, trueidx2 = trueidx2,
                        maxcancor = maxcancor,
                        Sigma1 = Sigma1, Sigma2 = Sigma2,
                        copula1 = "exp", copula2 = "cube",
                        muZ = mu,
                        type1 = "binary", type2 = "continuous",
                        c1 = rep(1, p1), c2 =  NULL
)
X1 <- simdata$X1; X2 <- simdata$X2; Sigma12_tt <- simdata$Sigma12
# Estimate latent correlation matrix with original method
R1_tt_org <- estR(X1, "binary", method = "original")
R2_tt_org <- estR(X2, "continuous", method = "original")
R12_tt_org <- estR(X1, type1 = "binary", X2, type2 = "continuous",
                              method = "original")$R12
# Estimate latent correlation matrix with original method
R1_tt_org <- estR(X1, "binary", method = "approx")
R2_tt_org <- estR(X2, "continuous", method = "approx")
R12_tt_org <- estR(X1, type1 = "binary", X2, type2 = "continuous",
                              method = "approx")$R12
```
